\documentclass[runningheads]{llncs}
\usepackage{longtable}

\begin{document}
\frontmatter
\setcounter{page}{5}

\chapter*{Preface}
\markboth{Preface}{Preface}
This volume contains the proceedings of the 28th International Conference on Coordination Models and Languages (COORDINATION 2026), held on June 8–12, 2026, in Urbino, Italy hosted by the
University of Urbino Carlo Bo., as part of the 21st International Federated
Conference on Distributed Computing Techniques (DisCoTec 2026).


Modern information systems rely increasingly on combining concurrent, distributed,
mobile, adaptive, reconfigurable, and heterogeneous components. New models, architectures, languages, and verification techniques are necessary to cope with the complexity induced by the demands of today’s software development. Coordination languages
have emerged as a successful approach, in that they provide abstractions that cleanly
separate behavior from communication, thereby increasing modularity, simplifying
reasoning, and ultimately enhancing software development. Building on the success of
the previous editions, COORDINATION 2026 provides a well-established forum for the
growing community of researchers interested in models, languages, architectures, and
implementation techniques for coordination.


COORDINATION 2026 solicited contributions in three different categories: (1) regular papers describing thorough and complete research results and experience reports;
(2) survey papers describing important results and success stories related to
the topics of COORDINATION; and (3) tool papers describing technological artifacts
in the scope of the research topics of COORDINATION, and requiring the submission
of the associated artifact via a separate procedure.


There were 28 paper submissions distributed over the different categories: 24 regular
papers, and 5 tool papers. The selection of the papers
was entrusted to the Programme Committee (PC), consisting of 32 members from 15
different countries (with gender balance F/M = 12/20). The selection of the papers was
done electronically, in two phases. In the first phase, which lasted three weeks, each
submission was single-blind reviewed by at least three PC members, in some cases with
the help of external reviewers. During the second phase, which lasted four days, the papers were thoroughly discussed. The decision to accept or reject a paper
was based not only on the reviews and scores but also on these in-depth discussions. In
the end, 13 papers were selected to be presented at the conference: 11 regular papers, and 2 tool papers. 


The authors of the submitted tool papers were requested to participate in the EAPLS
artifact badging, which instead was optional for all the other paper categories. There
were 13 artifact submissions. The 9 members of the Artifact Evaluation Committee
(AEC), chaired by Gianluca Aguzzi, awarded the Available and Reusable Badge to 3
artifacts, and the Available and Functional Badge to 4 artifacts.


We organized the accepted papers into five groups according to the topics they address:\begin{enumerate}
	\item Theoretical foundations and models for distributed systems: this group, containing three papers, focuses on formal reasoning of distributed system properties. The paper ``Timed Scenario Expressions and Realisability''  addresses the problem of realisability for the distributed system modeled by sets of sequential timed scenarios.  The paper ``HistMSO: a Logic for Reasoning on Consistency Models with MONA'' reasons on the consistency of replicated data systems modeled by sequences of data operations. The third paper ``Motif Refinement for the Hierarchical Control of Structured CPSs" introduces a refinement-based modeling and development approach for Cyber-physical systems which preserves control stability.  
	\item Distributed algorithms for collective adaptive systems: this group contains two papers which reformulate collective distributed algorithms within the aggregate programming framework with specific guarantees. The paper ``Aggregate Indoor Localisation" provides two existing established localisation algorithms to guarantees  adaptability and resilience while improving their performance. The paper ``A Self-Stabilizing Min-Max Consensus via Path-loop Detection" implements the min-max consensus algorithm as a
	reusable library for the Collektive DSL, which guarantees the self-stabilization property. 
	\item Coordination languages, frameworks, and patterns
	\item Formal verification in coordination languages: this group focuses on the static verification of two classes of coordination languages. The paper ``Proof of Delivery: Mechanized Mailbox Types" introduces a machine-checked foundation for the programming languages, specifically Pat, that incorporate mailbox type system to reason about the contents of mailboxes. The paper ``Deductive Verification of Legal Contracts" enables deductive verification of contracts specified by Stipula through their translation into Java programs annotated with Java 
	Modeling Language specifications preserving the dynamic
	features of the original programs – most notably the scheduling of events at future times.
	\item Simulation tools and methodologies for collective adaptive systems
\end{enumerate} 


As Programme Chairs, we actively contributed to the selection of the three keynote
speakers of DisCoTec 2026:\begin{itemize}
	\item Prof. Mehdi Dastani (Utrecht University, The Netherlands);
	\item Paolo Romano (NESC-ID, Portugal);
	\item Nathalie Bertrand (Inria, France).
\end{itemize}
We are most grateful to Prof. Dastani for accepting our invitation as the  COORDINATION-related keynote speaker. This volume includes the abstract of his keynote
talk: “Efficient Communication and Coordination in Multiagent Reinforcement Learning”.


As is traditional in DisCoTec, a joint session with the best papers from each main conference was organized. The best paper of COORDINATION 2026 was “Runtime adaptation as a programming pattern in service composition” authored by Carlos Gustavo Lopez Pombo, Pablo Montepagano and Emilio Tuosto. 


We are grateful to all the persons involved in COORDINATION 2026. In particular,
we thank the authors for their submissions, the attendees of the conference for their
participation, the PC members and external reviewers for their work in reviewing sub-
missions and participating in the discussions, the AEC chair Gianluca Aguzzi and
the AEC members for their effort in the evaluation of the artifacts, and the Steering Committee,
chaired by Michele Loreti, for their guidance and support. We are also grateful to the
Organizing Committee, chaired by Claudio Antares Mezzina, for their excellent job. We also thank
the providers of the EasyChair Conference Management System, which was of great
help in the submission and reviewing process and in the preparation of the proceedings.
We would also like to acknowledge the prompt and professional support from Springer,
which published these proceedings in printed and electronic volumes as part of their
LNCS book series.
\medskip

\begin{flushright}
\noindent April 2026\hfill
{[Roberto Casadei]}\\
{[Fatemeh Ghassemi]}
\end{flushright}


\chapter*{Organization}
\markboth{Organization}{Organization}
\setlength{\tabcolsep}{0pt}     % remove spacing between columns in the following long tables
\setlength{\LTpre}{0pt}         % remove extra vertical space before the long tables

%\section*{General Chair}
%\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
%Robert Bai          & University of Cambridge, UK
%\end{longtable}
%
%\section*{Program Committee Chairs}
%\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
%Vincent Garc\'{\i}a & Technische Universit\"{a}t Dresden, Germany\\
%Laurent Ho          & Dublin City University, Ireland\\
%Steffen Rao         & Carleton University, Canada\\
%\end{longtable}
%
%\section*{Steering Committee}
%\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
%Michael Agarwal     & KU Leuven, Belgium\\
%John Banerjee       & McMaster University, Canada\\
%Antonio Chan        & Harvard University, USA\\
%Stefano Guo         & Ritsumeikan University, Japan\\
%Mohamed Fang        & University of Porto, Portugal\\
%Alessandro Davis    & University of the Aegean, Greece\\
%Rafael Hoffmann     & Hanyang University, South Korea\\
%Mike L\'{o}pez      & Universidad de Granada, Spain\\
%Marek Keller        & University of Calabria, Italy\\
%Adrian Kwon         & Massachusetts Institute of Technology, USA\\
%Hans Meyer          & University of Kaiserslautern, Germany\\
%Nicholas Pinto      & Siemens AG, Germany\\
%Marcus Reddy        & ETH Zurich, Switzerland\\
%Sandra Ryu          & University of Bonn, Germany\\
%Patricia Silva      & Universidad Aut\'{o}noma de Madrid, Spain\\
%Edward  Smith       & King's College London, UK\\
%Tom\'{a}\v{s} Weber & University of Turku, Finland\\
%Gerhard Wu          & Masaryk University, Czech Republic\\
%Hiroki Yamamoto     & Bielefeld University, Hungary\\
%Werner Yoshida      & University of Catania, Italy\\
%\end{longtable}

\section*{Program Committee}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
	
S. Akshay &	IIT Bombay, India\\
Duncan Paul Attard & University of Malt, Malta \\
Giorgio Audrito &	University of Turin, Italy \\
Simon Bliudze &	INRIA, France \\
Valentina Castiglioni &  Eindhoven University of Technology, Netherlands \\
Mariangiola Dezani-Ciancaglini &	University of Turin, Italy\\
Cinzia Di Giusto &	Université Côte d’Azur, France \\
Francisco Ferreira & University of London, UK \\
Silvia Ghilezan & University of Novi Sad, Mathematical Institute SASA, Serbia\\
Susanne Graf & CNRS,VERIMAG, France\\
Ludovic Henrio & CNRS, France; IRIF \\
Thomas T. Hildebrandt & 	University of Copenhagen, Denmark \\
Ping Hou & University of Oxford, UK \\
Eva Kühn & Vienna University of Technology, Austria \\
Ivan Lanese & University of Bologna, Italy \\
Carlos López Pombo &  Universidad Nacional de Río Negro, Argentina; CONICET \\
Michele Loreti &	University of Camerino, Italy \\
Stefano Mariani & University of Modena and Reggio Emilia, Italy \\
Hernán C. Melgratti & University of Buenos Aires, Argentina \\
Fabrizio Montesi & University of Southern Denmark, Denmark \\
J. Garreth Morris & University of Iowa, USA\\
Rumyana Neykova & Brunel University London, UK \\
José Proença & University of Porto, Portugal \\
Violet Ka I Pun & Western Norway University of Applied Sciences, Norway \\
António Ravara & 	NOVA LINCS, Portugal \\
Marjan Sirjani & 	Mälardalen University, Sweden \\
Carolyn L. Talcott & 	SRI International, USA \\
Maurice H. ter Beek & ISTI-CNR, Italy \\
Francesco Tiezzi & University of Florence, Italy \\
Nils Timm & University of Pretoria, South Africa \\	
Emilio Tuosto & Gran Sasso Science Institute, Italy \\
Frank Valencia & École Polytechnique de Paris, France
\end{longtable}

\section*{Artefact Evaluation Committee}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
Davide Domini & University of Bologna, Italy \\
Nicolas Farabegoli & University of Bologna, Italy\\
Matteo Magnini & University of Luxembourg, Luxembourg\\
Mário Pereira & NOVA School of Science and Technology, Portugal\\
Marco Quadrini & University of Camerino, Italy\\
Corentin Reuther & University of Namur, Belgium \\
Lorenzo Rossi & University of Camerino, Italy \\
Gerard Tabone & University of Malta, Malta\\
Laura Voinea & University of Glasgow, UK
\end{longtable}

\section*{Additional Reviewers}
\begin{multicols}{2}
\noindent Nicola Del Giudice \\
Masoud Ebrahimi \\
Gabriele Genovese \\
Ramchandra Phawade \\
Jose Ignacio Requeno\\
Volker Stolz \\
\end{multicols}

\tableofcontents

\mainmatter

%Create dummy chapters for the sample ToC

%-----keynote
\setcounter{page}{1}
\addtocontents{toc}{\protect\section*{Keynote}}
\include{keynote}

%first heading
\setcounter{page}{1}
\addtocontents{toc}{\protect\section*{Theoretical foundations and models for distributed systems}}
\author{Neda Saeedloei and Feliks Kluzniak}
\title{Timed Scenario Expressions and Realisability}
\maketitle
\clearpage

\setcounter{page}{22}
\author{Isabelle Coget and Étienne Lozes}
\title{HistMSO: a Logic for Reasoning on Consistency Models with MONA}
\maketitle
\clearpage

\setcounter{page}{44}
\author{Simon Bliudze, Sophie Cerf and Olga Kouchnarenko}
\title{Motif Refinement for the Hierarchical Control of Structured CPSs}
\maketitle
\clearpage

\setcounter{page}{31}
\addtocontents{toc}{\protect\section*{Software Engineering and Applications for Coordinated Systems}}
\author{Still Another Author}
\title{A Paper About Something}
\maketitle
\clearpage

\setcounter{page}{41}
\author{Same Author}
\title{Last Paper of the Sample ToC}
\maketitle
\clearpage

\setcounter{page}{51}
\phantom{Author Index}
\addcontentsline{toc}{title}{\protect\textbf{Author Index}}
\clearpage

\setcounter{page}{31}
\addtocontents{toc}{\protect\section*{Coordination Platforms, Middleware, and Tools}}
\author{Still Another Author}
\title{A Paper About Something}
\maketitle
\clearpage
\end{document}
