\documentclass[runningheads]{llncs}
\usepackage{longtable}

\begin{document}
\frontmatter
\setcounter{page}{5}

\chapter*{Preface}
\markboth{Preface}{Preface}

It is the task of the volume editor to prepare the Organization pages and the
Preface and to structure the volume in comprehensible parts. The Preface
should give a good introduction to the volume and present the conference or
workshop, mentioning the venue, dates, and the focus of the event. Clear
information regarding the peer review process (single or double
blind\,/\,number of reviews per paper) and the numbers and types (full,
short, invited) of papers submitted and accepted should be given. If invited
papers are included in the main part of the manuscript, please include a
sentence or two in the Preface explaining how they were reviewed. For
contributions co-authored by any committee members, please explain in the
Preface how these were objectively handled during the review phase.

The Preface is usually one or two pages long and is followed by the
Organization pages. By default, for Springer Computer Science proceedings
volumes, the volume editors (the people named on the book cover) should be
the Program Committee Chairs, and they should also be the people who write
the Preface. In this way, the Program Committee Chairs take responsibility
for the scientific program of the conference, and for decisions on the
content of the Springer-published proceedings.

If you plan any other approach, i.e., if you plan to have volume editors who
are not the Program Committee chairs, please contact the Springer Editor to
explain and to seek approval. If approved, we ask you to include a sentence
about this in the Preface.

The names and roles of all of the Chairs should be included on the
Organization pages. When listing the Program Committee members, kindly just
state their names, their universities/affiliations and their countries and do
not include the departments or the institutes within the universities. The
names of the committee members should be ordered alphabetically, according to
surname, as shown on the Organization pages following on from this Preface.
Please also make sure that affiliations are spelt consistently.

At the end of the Organization pages, the sponsors may be listed and their
respective logos displayed. The Steering Committee members are usually listed
above the Program Committee members.

We prefer the use of American English in the Preface and we ask you to write
the Preface in the past tense, because it is mainly read after the event. The
Preface and the Organization pages will be copy-edited at our end, and any
corrections are submitted to the corresponding volume editor for approval.
Light technical copy-editing of the papers may also be performed.

The final Table of Contents is created by Springer from the title information
in the papers. However, we ask you to provide us with a preliminary Table of
Contents when you send the files. This should contain the titles of the
papers and the names of the authors in the order in which they are to appear
in the volume and should include topical section headings. The papers should
be grouped according to the topics they address and not according to the
sessions of the conference. Any session numbering will be deleted by our
typesetters. Please include all papers pertaining to one topic under one
single topical heading, even if there are/were two or three such sessions at
the conference. If the titles given in your preliminary Table of Contents
differ from those in the papers, then we take the paper titles to be the
correct ones and create the final Contents accordingly.

The Author Index is also generated at Springer, but you can help us present
the authors’ names in the correct way by submitting a list of authors who
have complex family names, particles, or suffixes, or do not use the Western
name order (i.e., given name(s); family name). It should be made quite clear,
which part of the name is the given name and which is the family name.

Please check that the names of the authors are written consistently
throughout the proceedings volume. If an author has contributed to more than
one paper, their name should be spelt and structured in an identical manner
in all papers as well as in the Table of Contents.

At the end of the preface, the month and year should be inserted on the
left-hand side and the names of the authors of the preface (i.e., the PC
Chairs) should be listed on the right-hand side. Please note that pages I--IV
(in front of the Preface) are prepared by Springer.
\medskip

\begin{flushright}
\noindent August 2025\hfill
{[name of volume editor]}\\
{[name of volume editor]}
\end{flushright}


\chapter*{Organization}
\markboth{Organization}{Organization}
\setlength{\tabcolsep}{0pt}     % remove spacing between columns in the following long tables
\setlength{\LTpre}{0pt}         % remove extra vertical space before the long tables

\section*{General Chair}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
Robert Bai          & University of Cambridge, UK
\end{longtable}

\section*{Program Committee Chairs}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
Vincent Garc\'{\i}a & Technische Universit\"{a}t Dresden, Germany\\
Laurent Ho          & Dublin City University, Ireland\\
Steffen Rao         & Carleton University, Canada\\
\end{longtable}

\section*{Steering Committee}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
Michael Agarwal     & KU Leuven, Belgium\\
John Banerjee       & McMaster University, Canada\\
Antonio Chan        & Harvard University, USA\\
Stefano Guo         & Ritsumeikan University, Japan\\
Mohamed Fang        & University of Porto, Portugal\\
Alessandro Davis    & University of the Aegean, Greece\\
Rafael Hoffmann     & Hanyang University, South Korea\\
Mike L\'{o}pez      & Universidad de Granada, Spain\\
Marek Keller        & University of Calabria, Italy\\
Adrian Kwon         & Massachusetts Institute of Technology, USA\\
Hans Meyer          & University of Kaiserslautern, Germany\\
Nicholas Pinto      & Siemens AG, Germany\\
Marcus Reddy        & ETH Zurich, Switzerland\\
Sandra Ryu          & University of Bonn, Germany\\
Patricia Silva      & Universidad Aut\'{o}noma de Madrid, Spain\\
Edward  Smith       & King's College London, UK\\
Tom\'{a}\v{s} Weber & University of Turku, Finland\\
Gerhard Wu          & Masaryk University, Czech Republic\\
Hiroki Yamamoto     & Bielefeld University, Hungary\\
Werner Yoshida      & University of Catania, Italy\\
\end{longtable}

\section*{Program Committee}
\begin{longtable}{p{0.3\textwidth}p{0.7\textwidth}}
Thomas Ahmad        & University of Minho, Portugal\\
Paul Bao            & Nagoya University, Japan\\
Roberto Chu         & University of Melbourne, Australia\\
Ruth de Oliveira    & University of Ljubljana, Slovenia\\
Paolo Feng          & University of South Australia, Australia\\
Jos\'{e} Geng       & Shandong University, China\\
Johannes Gupta      & University of Electronic Science and Technology, China\\
Juan Hong           & Polish Academy of Sciences, Poland\\
Ricardo Islam       & University of Oslo, Norway\\
Pavel Jiao          & University of Oxford, UK\\
Julien Khan         & RWTH Aachen, Germany\\
Ahmed Lai           & Helsinki University of Technology, Finland\\
Ulrich Liao         & University of North Carolina, USA\\
Jason Lu            & Jadavpur University, India\\
Samuel Ng           & University of Granada, Spain\\
Daniele Park        & University of Lisbon, Portugal\\
Nick Ren            & University of California, USA\\
Alexandra Saha      & Purdue University, USA\\
Bernard Shang       & Beijing University of Technology, China\\
Jeff Simon          & Swinburne University of Technology, Australia\\
Harald Tan          & Anna University, India\\
Oscar Tran          & Johannes Kepler University, Austria\\
Kenji Wei           & National University of Defense Technology, China\\
Axel Xia            & Monash University, Australia\\
Karl You            & Lund University, Sweden\\
Manfred Zhao        & University of Duisburg-Essen, Germany
\end{longtable}

\section*{Additional Reviewers}
\begin{multicols}{2}
\noindent Peter Ahn\\
Martin Ali\\
Christian Almeida\\
Marco Bauer\\
Andrea Becker\\
Mark Brown\\
Richard Chen\\
Marc Cheng\\
Eric Cheung\\
Chris Chung\\
Jonathan Cohen\\
Nicolas Costa\\
Florian De Souza\\
Tobias Deng\\
Alberto Ding\\
Matthew Fernandes\\
Pedro Fern\'{a}ndez\\
Luca Ferreira\\
Jorge Ghosh\\
Alex Gomes\\
Vladimir G\'{o}mez\\
Laura Ha\\
Bruno Hall\\
William Han\\
Zhang Hou\\
Christophe Hsieh\\
Michel Hsu\\
Victor Ito\\
Jens Jain\\
Yang Jang\\
Igor Jin\\
Fabio Jing\\
Gabriel Johnson\\
Philipp Kim\\
Sven King\\
J\"{u}rgen Klein\\
Sara Lam\\
Hiroshi Lau\\
Joseph Le\\
Olga Lim\\
Guillaume Lima\\
Claudia Lin\\
Ralf Ludwig\\
Fr\'{e}d\'{e}ric Luo\\
Charles Ma\\
Anton Miller\\
Alejandro Min\\
Ioannis Mishra\\
Andrzej Nguyen\\
Claudio Ni\\
Simone Nie\\
Petr Patel\\
Davide Paul\\
Johan Pei\\
Dmitry Qian\\
Silvia Qiao\\
Holger Qin\\
Yuki Ribeiro\\
Takeshi Roberts\\
Julia Robinson\\
Walter S\'{a}nchez\\
Marcel Santos\\
Jean Sato\\
Alexey Shao\\
Dimitris Sharma\\
Mikhail Shen\\
Nikolaos Singh\\
Edward Smith\\
Jonas Son\\
Catherine Tanaka\\
Masahiro Tang\\
Sylvain Tao\\
Timo Tsai\\
Ren\'{e} Tseng\\
Omar Tu\\
Lorenzo Wei{\ss}\\
Gilles Wen\\
Mauro Weng\\
Christina Xiang\\
Jacques Xiao\\
Tony Xie\\
Grzegorz Yang\\
Ryan Yao\\
Elizabeth Ye\\
Guido Young\\
Antoine Yu\\
Andr\'{e}s Yuan\\
Gregory Zheng\\
Aaron Zhong\\
Anders Zhou
\end{multicols}

\tableofcontents

\mainmatter

%Create dummy chapters for the sample ToC

\setcounter{page}{1}
\addtocontents{toc}{\protect\section*{First Topic}}
\author{One Author and Another Author}
\title{This Is a Sample Title}
\maketitle
\clearpage

\setcounter{page}{11}
\author{Some Author}
\title{This Is Another Contribution Title}
\maketitle
\clearpage

\setcounter{page}{21}
\author{First Author and Second Author}
\title{A Third Paper Title}
\maketitle
\clearpage

\setcounter{page}{31}
\addtocontents{toc}{\protect\section*{Second Topic}}
\author{Still Another Author}
\title{A Paper About Something}
\maketitle
\clearpage

\setcounter{page}{41}
\author{Same Author}
\title{Last Paper of the Sample ToC}
\maketitle
\clearpage

\setcounter{page}{51}
\phantom{Author Index}
\addcontentsline{toc}{title}{\protect\textbf{Author Index}}
\clearpage
\end{document}
