\section{The BLPy coordination language}
\label{sect-BLPy}

\subsection{The Bach Coordination Language}

Bach is a Linda-like coordination language developed at the
University of Namur~\cite{jacquet2007coordinating}. It has been conceived in a
constraint logic programming flavor. As a result, Linda tuples are
replaced by ground logic programming terms, also called si-terms, the
shared space is referenced as the store, and Linda primitives are
reformulated as the following primitives: $\mathit{tell(t)}$ to place an
occurrence of $t$ on the store, $\mathit{ask(t)}$ to verify that $t$
is present on the store, $\mathit{get(t)}$ to proceed similarly but by
consuming one occurrence of $t$, and $\mathit{nask(t)}$ to check that
$t$ is absent from the store. Note that the last three primitives
suspend until their associated condition is met. 

Matching is operated in Bach as syntactic equality. We
shall soften it in this paper by allowing a partially specified
si-term to match a more complete si-term.  This will be very useful
later to get values from non specified arguments.  The precise
definition is as follows.

\begin{definition}
Let $t=f(t_1,\cdots,t_m)$ and $u=g(u_1, \cdots, u_n)$ be two
si-terms. Then $t$ matches $u$ iff, on the one hand, their functors
are the same ie $f=g$ and, on the other hand, the list
$[t_1,\cdots,t_m]$ of arguments of $t$ is a prefix of the list
$[u_1,\cdots,u_n]$ of arguments of $u$. This is subsequently denoted by
$t \triangleleft u$. On the point of terminology, in that case, it is also said that 
%$u$ is matched by $u$.
$u$ is matched by $t$.
\end{definition}

Primitives are composed in Bach to form more complex instructions,
called agents, by using traditional composition operators from
concurrency theory: sequential composition, parallel composition and
non-deterministic choice. It is worth noting, however, that we shall only use the two first subsequently.





\subsection{The BLPy implementation}

% OK pour prop de Manel
%For the purpose of exploiting federated inductive logic programming using Bach, an implementation has been developed in Python. It is based on sockets and  thus is mainly organized around, on the one hand, a server, managing the store and the actions of primitives, and, on the other hand, clients, interacting with the store by means of the primitives. %The use of sockets is quite classical. 
%The use of sockets is standard. Taking 127.0.0.1 as %local IP address and 8000 as port number,
%the local IP address 127.0.0.1 and port 8000, the server and client exchange messages of at most 1024 bytes, in loops basically consisting of sending and receiving these messages. That being said, the core part is threefold : it consists in defining a grammar to parse the messages, in implementing the store, and in designing an interpreter. As regards the grammar, our work is based on the parsimonious module, which provides fast arbitrary-lookahead parsers written in pure Python. The store is implemented by using three data structures:

%------------  Proposition Manel ------------------%

For the purpose of exploiting federated inductive logic programming using Bach, an implementation has been developed in Python. It is based on sockets and organized around, on the one hand, a server, managing the store and the actions of primitives, and, on the other hand, clients interacting with the store. The core part is threefold: it consists in defining a grammar to parse the messages, in implementing the store, and in designing an interpreter. As regards the grammar, our work is based on the \texttt{parsimonious} module, which provides fast arbitrary-lookahead parsers written in pure Python. The store is implemented using three data structures:

\begin{sloppypar} 
\begin{itemize}

\item a dictionary, named \texttt{theStore}, which represents the real contents of the store. For each functor it contains as a key it associates a dictionary mapping si-terms formed with this functor to integers, denoting the numbers of occurrences of the considered si-terms currently on the store;

\item a dictionary, named \texttt{theWaitingList}, that registers process identifiers waiting for the presence of si-terms. Technically, it associates with a functor name a list of triples, composed of a process identifier, the si-term on which the process is suspended and the destructive nature of the operation (according to the fact that the process is suspended on a ask or a get request);

\item a dictionary, named \texttt{theNWaitingList}, that registers process identifiers waiting for the absence of si-terms. Its structure is similar to that of \texttt{theWaitingList} except that processes are there waiting for the absence of the corresponding si-terms and that no destructive feature is recorded.
\end{itemize} 
\end{sloppypar}

\noindent
That being said, the implementation of the primitives is as follows:

\begin{itemize}

\item the execution of $tell(t)$ consists in checking whether the
  functor of $t$, say $f$, is registered in \texttt{theStore}. If this
  is the case and if $t$ already appears as a key of
  \texttt{theStore[f]} then the number of associated occurrences is
  increased by one. If $f$ is a key of \texttt{theStore} but $t$ is
  not a key of \texttt{theStore[f]} then a new entry is added for $t$
  with 1 as associated integer. If $f$ is not a key of
  \texttt{theStore} then it is introduced in \texttt{theStore} keys
  with a mapping from $t$ to $1$ as associated value. In all the
  cases, the processes of \texttt{theWaitingList} are looked at to see
  whether processes can be waken up.

\item the execution of $ask(t)$ is dual. It succeeds in case the
  functor $f$ of $t$ is a key of \texttt{theStore} and if there is a
  si-term matching $t$ in \texttt{theStore[f]}. If this is not the
  case then the identifier of the process running $ask(t)$ together
  with $t$ is added to \texttt{theWaitingList}.

\item the execution of $get(t)$ is similar but consists of removing an
  occurrence of a matching $u$ is case such a si-term is present. In a
  dual perspective as $tell(t)$ processes of \texttt{theNWaitingList}
  are inspected to determine whether they can be waken up.

\item finally the execution of $nask(t)$ is dual to that of $ask(t)$, asking for the absence of si-terms $u$ matched by $t$ and resulting in the process identifier being added to \texttt{theNWaitingList} together with $t$ in case of suspension.
  
\end{itemize}

Sequential and parallel compositions are not implemented directly, but rather 
%are treated at the Python level.
are handled at the Python level.

As a proof of concept, a Python package, named BLPy, has been developed. It can be downloaded at the address mentioned in~\cite{BLPy}. 

%In addition to the necessary files for handling the grammar, for executing the interpreter and for implementing the store and the primitives acting on it, the BLPy module is made up of two main files. On the one hand, a file, called \texttt{blsrv.py}, allows to execute the server. It basically creates an instance of a store, of a parser, and of an interpreter. On that basis, it then listens for the primitives transmitted by the clients, executes them, and sends the results of the execution to the clients. On the other hand, a file, named \texttt{blcli.py}, allows to execute a client. It basically consists of a REPL through which the user can introduce primitives to be executed and receive results from the server. 

% Figure~\ref{blpy-test} shows the execution of three terminals, with the top left one launching the server and the two others clients. 

% \begin{figure}[t]
% Figure to be given by Manel
% \caption{Sample of execution in BLPy
% \label{blpy-test}}
% \end{figure}











